### 2017-03-21 doc/manuscript.Rmd
Added Tony's edits to manuscript

Title: Composition of fatty acids within the serum triacylglycerol fraction and 
the association with the pathogenesis of diabetes

Author: Luke W. Johnston, MSc (1); Zhen Liu, PhD (1); Ravi Retnakaran, MD
(3,4); Bernard Zinman, MD (3,4); Adria Giacca, PhD (5); Stewart B. Harris, MD (2); 
Richard P. Bazinet, PhD (1); and Anthony J. Hanley, PhD (1,6)

@@

Hypertriglyceridemia is an extensively studied and well known factor involved in
the dysregulation of metabolic function and subsequent negative health outcomes [@Chehade2013a].
It is a component of the metabolic syndrome [@Alberti2009] and a strong risk factor for
cardiovascular disease [@DAgostino2004a], in addition to it's role in metabolic
dysfunction [@Verges2015a; @Kawano2013a]. The measurement of circulating 
triacylglcerides (TAG) is commonly done during routine clinical visits as part
of health assessment. However, clinically measured TAG is limited as it
represents all the types of fatty acids within the TAG fraction as a single
measure. Given the increasing appreciation for the role of the specific fatty
acid composition on health outcomes, for instance with differences in
associations of fatty acids within the PL lipid fraction [@Forouhi2014a; @Ma2015a] 
including an analysis from the PROMISE cohort [@Johnston2016a], there
are relatively few studies that have explored this area of research in the TAG
fraction.

The interaction between TAG and insulin sensitivity is complex and has 
components of a feedback system [@Verges2015a]. Greater resistance to insulin in
both the liver and muscle may result in greater production of TAG and secretion
of lipoproteins that transport TAG [@Yu2012a]. Likewise, greater TAG may
contribute to metabolic dysfunction and lipotoxicity in various tissues, such as
the beta-cells, and thus continuing the cycle [@Verges2015a]. Given the
complexity and temporal nature of the relationship, long term studies with
multiple data collections are paramount to better understanding the underlying
biology and subsequent risk.

While several prospective studies have documented a prospective association of 
hypertriglyceridemia with incident type 2 diabetes 
[@Chien2008a; @DAgostino2004a; @Schulze2009a], there are limited longitudinal
studies [@Rhee2011a; @Lankinen2015a] that have examined the relationship between
TAG and it's composition on the pathophysiological factors underlying diabetes, 
particularly in regard to beta-cell function. Our objective was to examine the 
longitudinal role of the specific composition of the serum TAG fraction on 
OGTT-derived measures of insulin sensitivity and beta-cell function compared to 
clinically measured TAG in an Canadian population at risk for diabetes.

@@

Recruitment for the baseline visit of the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
cohort took place between 2004-2006 in London and Toronto, 
Canada. Individuals were selected to participant if they met the eligibility 
criteria of having one or more risk factors for type 2 diabetes mellitus, 
including obesity, hypertension, family history of diabetes, and/or a history of
gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
attended the baseline visit. Subsequent examinations occurred every three years,
with three examination visits completed to date (2004-2006, 2007-2009, and
2010-2013). The current study
used data on participants who did not have diabetes at baseline, who
returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477; see the CONSORT diagram in 
`r cite_sf('consort')`). Metabolic characterization, anthropometric
measurements, and questionnaires on lifestyle and sociodemographics were
administered at each examination visit. Research ethics approval was obtained from
Mount Sinai Hospital and the University of Western Ontario, and all participants
provided written informed consent. Data collection methods were standardized
across the 2 centres and research nurses were centrally trained.

@@

After 8-12 hours of overnight fasting, participants completed a 75g oral glucose tolerance test (OGTT) at 
each examination visit, with blood samples taken at fasting, 30 min, and 2 hr 
post-glucose load. Samples were subsequently processed and frozen at -70°C.
Alanine aminotransferase (ALT) was measured using standard laboratory 
procedures. Cholesterol, HDL, and clinically-measured TAG 
were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
ON). Both insulin and glucose were
measured from OGTT blood samples at fasting, 30 minute,
and 2 hour time points. Specific insulin was measured with the Elecsys 1010 
(Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and 
electrochemiluminescence immunoassay, which shows 0.05% cross-reactivity to 
intact human pro-insulin and the Des 31,32 circulating split form (Linco Res. 
Inc) and has a coefficient of variation (CV) of 9.3%. Glucose was determined 
using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
detection range of 0.11 to 41.6 mmol/L and an inter-assay %CV of <1.1% and an
intra-assay %CV of < 1.9%.  All assays were performed at the Banting and Best
Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
impaired glucose tolerance (IGT), and diabetes were categorized using the 2006
WHO criteria [@WHO2006].

@@

In analyses using concentration values, nearly all TAGFA had a strong
negative association on HOMA2-%S and ISI (estimates ranging from
`r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
few had strong negative associations with IGI/IR and ISSI-2 (estimates ranging 
from `r est1$outcome_b$rng[11]` and `r est1$outcome_b$rng[15]`, respectively).
In analyses using TAGFA mol% values, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
associations with HOMA2-%S and ISI (between `r est1$outcome_b$rng[1]` and 
`r est1$outcome_b$rng[5]`, respectively, lower insulin sensitivity for every SD
increase in the TAGFA), while several more TAGFA had positive associations with
HOMA2-%S and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
and 22:5n-3) predicting between `r est1$outcome_b$rng[2]` and 
`r est1$outcome_b$rng[6]`%, respectively, higher insulin sensitivity for every
SD increase in the TAGFA. One TAGFA, 20:2n-6, had a very strong positive
association with the insulin sensitivity measures, with a `r est1$b_ci$est[1]`
to `r est1$b_ci$est[37]`% higher insulin sensitivity for every SD increase. 
Both clinically-measured TAG and total TAGFA concentration had very strong
negative associations with all outcome variables.

@@

The PLS analysis corroborated the findings from the GEE models. 
The PLS results conditioned on insulin sensitivity as the outcome showed a
clustering of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7 
(`r cite_f('pls')`). These TAGFA loaded strongly and negatively on HOMA2-%S and ISI,
suggesting this cluster of TAGFA tracks
together with lower insulin sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3,
and 22:6n-3 loaded positively on both insulin sensitivity measures. No other
TAGFA loaded strongly. The PLS model for insulin sensitivity had good predictive
ability, with a high correlation between the predicted outcome values against
the observed values (HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).

@@

To our knowledge, no longitudinal study to date has examined the role of the
composition of the TAGFA fraction on detailed OGTT-derived metabolic measures 
Two large prospective studies have been published that similarly examined TAGFA 
composition and diabetes outcomes. Rhee *et al* presented a nested case-control
analysis (n=189 cases and n=189 controls) within the Framingham offspring cohort
[@Rhee2011a], which found that subjects with a TAGFA composition characterized
by a lower carbon chain and fewer double bonds (e.g. 14:0, 16:0) had a higher
risk for diabetes after 12-years while those with a profile characterized by
higher carbon chain and more double bond TAGFA had a lower risk for diabetes.
A similar pattern of TAGFA was also associated with HOMA-IR cross-sectionally
at the baseline visit. In addition, Lankinen *et al* reported on
a prospective cohort of males in Finland [@Lankinen2015a], for which 
TAGFA data were available for 831 participants after 6-years of follow-up. In
their cohort, OGTT data was only available at the 6-year visit.
They cross-sectionally found that most saturated fatty acids had negative 
associations with insulin sensitivity and beta-cell function while
linoleic acid (18:2n-6), docosapentaenoic acid (22:5n-3), eicosapentaenoic acid
(20:5n-3), and arachidonic acid (20:4n-6) had positive associations with insulin
sensitivity. The magnitude of the associations were larger in the insulin
sensitivity results compared to the beta-cell function results, similar to what
we observed. Our study extends these findings by using multiple measurements
of metabolic function and as well as multivariate statistical approaches that
allowed us to identify clusters of TAGFA. In another study of a much
smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
for the 18:1n-9 finding.

@@

The direction of association between TAGFA and insulin sensitivity is unclear 
from cross-sectional studies due to the physiological feedback mechanisms 
involved. For example, while greater DNL may promote muscle insulin resistance, 
the reverse may also be true [@Flannery2012a]. Higher insulin resistance may
encourage greater DNL to handle the higher blood glucose. To illustrate this, in
a weight loss intervention trial (n=19 with TAGFA data), participants who lost
weight over 33 weeks showed higher insulin sensitivity and lower TAGFA
composition indicative of lower DNL (less 14:0, 14:1, 16:0, 16:1, etc)
[@Schwab2008a]. Given the complex biological mechanisms and feedback loops
involved, disentangling whether insulin sensitivity influences TAGFA more
strongly than TAGFA influencing insulin sensitivity will require more
complicated research designs and analyses. While further research will need to
confirm this, the lack of a time interaction we observed in addition to the
timing of our measures likely suggests that it is the TAGFA predicting insulin
resistance.

>>>>

Title: Specific clusters of fatty acids within the serum triacylglycerol fraction 
associate with the pathogenesis of type 2 diabetes

Author: Luke W. Johnston, MSc (1); Zhen Liu, PhD (1); Ravi Retnakaran, MD
(3,4); Bernard Zinman, MD (3,4); Adria Giacca, PhD (5); Stewart B. Harris, MD (2); 
Richard P. Bazinet, PhD (1); and Anthony J. Hanley, PhD (1,3,6)

@@

Hypertriglyceridemia is an extensively studied and well described disorder in
the dysregulation of metabolic function and subsequent negative health outcomes [@Chehade2013a].
It is a strong risk factor for cardiovascular disease [@DAgostino2004a], in addition to other metabolic disorders such as non-is involved in metabolic dysfunction [@Verges2015a; @Kawano2013a; @Alberti2009a]. 
Circulating triacylglceride (TAG) concentration is commonly measured during
routine clinical assessment using enzymatic methods.
However, clinically measured TAG is limited as it
represents the full fatty acid spectrum within the TAG fraction as a summary
measure. There is increasing appreciation for the importance of specific fatty 
acid composition profiles in different plasma fractions on various health
outcomes [@Forouhi2014a; @Ma2015a; @Johnston2016a], however there are relatively
few studies that have explored the impact of the fatty acid composition in the TAG fraction [@Rhee2011a; @Lankinen2015a].

The interaction between TAG and insulin sensitivity is complex and involves 
components of a feedback system [@Verges2015a]. Greater resistance to insulin in
both the liver and muscle may result in greater production of TAG and secretion
of lipoproteins that transport TAG [@Yu2012a]. Likewise, greater TAG may
contribute to metabolic dysfunction and lipotoxicity in various tissues, such as
the beta-cells, and thus continuing the cycle [@Verges2015a]. Given the
complexity and temporal nature of the relationship, long term studies with
multiple data collection time points are paramount to better understanding the underlying
biology and subsequent risk.

While several studies have documented prospective associations of 
hypertriglyceridemia with incident type 2 diabetes 
[@Chien2008a; @DAgostino2004a; @Schulze2009a], only a limited number of longitudinal
studies [@Rhee2011a; @Lankinen2015a] have examined the relationship between
TAG and its composition on the pathophysiological factors underlying diabetes, 
particularly beta-cell function. Our objective was to examine the 
longitudinal role of the specific composition of the serum TAG fraction on 
OGTT-derived measures of insulin sensitivity and beta-cell function compared to 
clinically measured TAG in a Canadian population at risk for diabetes.

@@

Recruitment for the baseline visit of the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
cohort took place between 2004-2006 in London and Toronto, 
Canada. Individuals were selected to participant if they met the eligibility 
criteria of having one or more risk factors for type 2 diabetes mellitus, 
including obesity, hypertension, family history of diabetes, and/or a history of
gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
attended the baseline visit. Subsequent examinations occurred every three years,
with data from three examination visits available for the present analysis
(2004-2006, 2007-2009, and 2010-2013). The current study
used data on participants who did not have diabetes at baseline, who
returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477; see the CONSORT diagram in 
`r cite_sf('consort')`). Metabolic characterization, anthropometric
measurements, and questionnaires on lifestyle and sociodemographics were
administered at each examination visit. Research ethics approval was obtained from
Mount Sinai Hospital and the University of Western Ontario, and all participants
provided written informed consent. Data collection methods were standardized
across the 2 centres and research nurses were centrally trained.

@@

After 8-12 hours of overnight fasting, participants completed a 75g oral glucose tolerance test (OGTT) at 
each examination visit, with blood samples taken at fasting, 30 min, and 2 hr 
post-glucose load. Samples were subsequently processed and frozen at -70°C.
Alanine aminotransferase (ALT) was measured using standard laboratory 
procedures. Cholesterol, HDL, and clinically-measured TAG 
were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
ON). Both insulin and glucose were
measured from OGTT blood samples at fasting, 30 minute,
and 2 hour time points. Specific insulin was measured with the Elecsys 1010 
(Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and 
electrochemiluminescence immunoassay, which shows 0.05% cross-reactivity to 
intact human pro-insulin and the Des 31,32 circulating split form (Linco Res. 
Inc) and has a coefficient of variation (CV) of 9.3%. Glucose was determined 
using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
detection range of 0.11 to 41.6 mmol/L and an inter-assay CV of <1.1% and an
intra-assay CV of < 1.9%.  All assays were performed at the Banting and Best
Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
impaired glucose tolerance (IGT), and diabetes were categorized using the 2006
WHO criteria [@WHO2006].

@@

In analyses using concentration values, nearly all TAGFA had a strong
negative association on HOMA2-%S and ISI (estimates of percent different ranging from
`r est1$outcome_b$rng[3]` and `r est1$outcome_b$rng[7]`, respectively), and a
few had strong negative associations with IGI/IR and ISSI-2 (estimates ranging 
from `r est1$outcome_b$rng[11]` and `r est1$outcome_b$rng[15]`, respectively).
In analyses using TAGFA mol% values, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7) had negative
associations with HOMA2-%S and ISI (between `r est1$outcome_b$rng[1]` and 
`r est1$outcome_b$rng[5]`, respectively, lower insulin sensitivity for every SD
increase in the TAGFA), while several more TAGFA had positive associations with
HOMA2-%S and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9, 18:2n-6, 20:2n-6, 20:4n-6,
and 22:5n-3) predicting between `r est1$outcome_b$rng[2]` and 
`r est1$outcome_b$rng[6]`%, respectively, higher insulin sensitivity for every
SD increase in the TAGFA. One TAGFA, 20:2n-6, had a very strong positive
association with the insulin sensitivity measures, with a `r est1$b_ci$est[1]`
to `r est1$b_ci$est[37]`% higher insulin sensitivity for every SD increase. 
Both clinically-measured TAG and total TAGFA concentration had very strong
negative associations with all outcome variables.

@@

The PLS analysis corroborated the findings from the GEE models. 
The PLS results conditioned on insulin sensitivity as the outcome showed a
clustering of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7 as mol%
(`r cite_f('pls')`). These TAGFA loaded strongly and negatively on HOMA2-%S and ISI,
suggesting this cluster of TAGFA tracks
together with lower insulin sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3,
and 22:6n-3 loaded positively on both insulin sensitivity measures. No other
TAGFA loaded strongly. The PLS model for insulin sensitivity had good predictive
ability, with a high correlation between the predicted outcome values against
the observed values (HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).

@@

To our knowledge, no longitudinal study to date has examined the role of the
composition of the TAGFA fraction on detailed OGTT-derived metabolic measures 
Two large prospective studies have been published that similarly examined TAGFA 
composition and diabetes outcomes. Rhee *et al* presented a nested case-control
analysis (n=189 cases and n=189 controls) within the Framingham offspring cohort
[@Rhee2011a], which found that subjects with a TAGFA composition characterized
by a lower carbon chain and fewer double bonds (e.g. 14:0, 16:0) had a higher
risk for diabetes after 12-years while those with a profile characterized by
higher carbon chain and more double bond TAGFA had a lower risk for diabetes.
A similar pattern of TAGFA was also associated with HOMA-IR cross-sectionally
at the baseline visit. In addition, Lankinen *et al* reported on
a prospective cohort of males in Finland [@Lankinen2015a], for which 
TAGFA data were available for 831 participants after 6-years of follow-up. In
their cohort, OGTT data was only available at the 6-year visit.
They cross-sectionally found that most saturated fatty acids had negative 
associations with insulin sensitivity and beta-cell function while
linoleic acid (18:2n-6), docosapentaenoic acid (22:5n-3), eicosapentaenoic acid
(20:5n-3), and arachidonic acid (20:4n-6) had positive associations with insulin
sensitivity. The magnitude of the associations were larger in the insulin
sensitivity results compared to the beta-cell function results, similar to what
we observed. Our study extends these findings by using multiple measurements
of metabolic function and as well as multivariate statistical approaches that
allowed us to identify clusters of TAGFA. In another study of a much
smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, findings which were largely 
similar to the present analysis.

@@

The direction of association between TAGFA and insulin sensitivity is unclear 
from previous cross-sectional studies due to the physiological feedback mechanisms 
involved. For example, while greater DNL may promote muscle insulin resistance, 
the reverse may also be true [@Flannery2012a]. Higher insulin resistance may
encourage greater DNL to handle the higher blood glucose. To illustrate this, in
a weight loss intervention trial (n=19 with TAGFA data), participants who lost
weight over 33 weeks showed higher insulin sensitivity and lower TAGFA
composition indicative of lower DNL (less 14:0, 14:1, 16:0, 16:1, etc)
[@Schwab2008a]. Given the complex biological mechanisms and feedback loops
involved, disentangling whether insulin sensitivity influences TAGFA more
strongly than TAGFA influencing insulin sensitivity will require more
complicated research designs and analyses. While further research will need to
confirm this, the lack of a time interaction we observed in addition to the
timing of our measures likely suggests that it is the TAGFA predicting insulin
resistance.
